Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-03-27 manuscript/Chapter2-manuscript.Rmd
update MS with new references style

---
title: "Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame"
output: 
  word_document:
    reference_docx: Chapter2-styles.docx
    fig_caption: yes
    pandoc_args: ["-Fpandoc-crossref"] # to have eq. numbers
bibliography: references.bib
csl: ecology.csl
---

@@

We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used Gaussian priors for the game duration ($N(0,\ 1)$), the intercept ($N(0,\ 0.5)$), and the precision paremeter ($N(2,\ 0.5)$). We employed the default Student t priors for the smoothing parameter ($t(3,\ 0,\ 2.5)$). In all models, we specified four MCMC chains to estimate the parameters, with 1000 posterior samples for each parameter. To do so, we ran 2500 iterations for the first model with a thinning set to eight, and 1500 iterations with a thinning set to four for the second and third models; burn-in was set to 500 iterations in each model. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021]. After fitting the three models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

We parameterized equations (1) and (2) with a Gaussian distribution family (predator and prey speed) and used a beta-binomial distribution family for equation (3) (predator hunting success). We used Gaussian priors for the prey rank ($N(2.5,\ 1)$) and game duration ($N(10,\ 2)$) on the mean part of the model, and the default priors in "brms" for the dispersion part. We used the default Student t priors in "brms" for the intercepts on the mean and dispersion part of the equation. For hunting success, we applied a Gaussian prior on the precision parameter ($N(2,\ 1)$). For every predicted variable, we applied a half-Gaussian on the random effects for the mean part of the model ($N(0,\ 1)$), and the default "brms" Student t prior for the dispersion part of the model. We parametrized the model to run four MCMC chains with 1000 posterior samples for each parameter by completing 2500 iterations having a thinning set to 8, with the first 500 iterations used as warmups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the model (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We additionally report variance partitioning and coefficients of variations in the supporting information.

@@

Our results confirm that individual predators display substantial differences in the development of their expertise, as the best model included group-level smoothers exclusively (Table I). For instance, some predators steadily increased (Figure 1A-S1A), while others decreased (Figure 1A-S1B), their hunting success with greater experience, and others whose success did not change with experience (Figure 1A). Another pattern that emerged was that some individuals optimized their success at an experience level that was below what we considered "advanced" (i.e. >300 matches played).

@@

At the population level, we predicted that predators would become either increasingly flexible or specialized with more experience. However, our results do not support this prediction. Indeed, the observed pattern is more complex than expected given that predators became more flexible with experience at intermediate levels of experience (i.e. intercept of sigma for predator speed is larger for intermediates than novices; Table 2 and Figure S4), but then return to novice-levels with advanced experience (i.e. intercept of sigma for predator speed is smaller for advanced players than intermediates; Table 2 and Figure S4). Thus, our analyses suggest that population-level behavioural specialization changes nonlinearly (quadratic) with experience rather than linearly.

@@

At the individual level, we detected differences in the mean movement speed among predators (i.e. the standard deviation of the mean speed for the predator ID; Table 3). These differences largely remained stable to the intermediate stage and then increased at the advanced stage to a level higher than when predators were novice (Table 3 and Figure S4). For the dispersion part of the model, predators displayed considerable differences in their degree of hunting specialization, which also increased with experience (i.e. the standard deviation of sigma for the predator ID increases with experience; Table 3 and Figure S4). Thus, the predator population is composed of both specialist and flexible individuals, and they further distinguish themselves as specialist/flexible hunters as they gain experience (Figure 2).

@@

As we predicted, predators differed in the mean speed of the prey that they encountered. Thus, some individuals encountered prey that were on average slower, while others encountered prey that where on average faster throughout the study period (i.e. the standard deviation of the mean prey speed for the predator ID; Table 3). These differences were greater when predators reached an advanced level of hunting experience (i.e. the standard deviation of the mean prey speed for the predator ID was higher at the advanced stage; Table 3 and Figure S4). In addition, individuals experienced largely distinct degrees of variability among their encounters with the prey. Hence, some predators experienced stable/predictable encounters with their prey, while others experienced unstable/unpredictable encounters with their prey throughout the study period (i.e. the standard deviation of sigma prey speed for the predator ID; Table 3).

@@

At all experience levels, predators that were on average faster (mean) also had a lower intra-individual variability (IIV) across their matches (Figure 3 and Table S1). Thus, faster predators specialized at travelling fast, while slower predators were more flexible in their movement through time. In general, predators that encountered prey that were on average faster (mean) tended to be faster (mean) (Figure 3 and Table S1). At the three experience levels, there were only weak correlations between predator IIV and mean prey speed (Figure 3 and Table S1). However, predators that experienced more predictable encounters with their prey (IIV) specialized in a faster hunting tactic (IIV) (Figure 3 and Table S1). For all the results described above, the correlations strengthen as predators gain experience (Figure 3 and Table S1).

The foraging strategies of predators, that is, the correlations between the mean and IIV of speed, remained structurally stable among the levels of experience (Figure 3 inside the black squares and Table S1). For example, novice predators that tended to be faster maintained this strategy as they gained experience (Figure 3 inside the black squares and Table S1). Overall, we detected the same patterns as those from the within-experience state correlations described above. The correlations between trait values were also stronger among novice-intermediate and intermediate-advanced states than between novice-advanced states, indicating that the hunting strategies of predators were more similar among experience levels closer in time (Figure 3 inside the black squares and Table S1).

@@

We did not detect strong correlations between the predators' mean speed and their hunting success (Figure 3 and Table S1), nor between the predators' IIV in speed and their hunting success (Figure 3 and Table S1), suggesting that slower/flexible and faster/specialist predators were both equally successful. However, when predators were novice, there was a strong negative correlation between the mean speed of the prey that they encountered and their hunting success (Figure 3 and Table S1). Hence, novice individuals that encountered faster prey were less successful. This relationship was decoupled at the later experience levels, such that encounters with faster prey at intermediate and advanced stages were not as strongly correlated with hunting success anymore (Figure 3 and Table S1). Thus, the consequences of hunting faster prey were reduced at greater experience levels.

@@

Empirical research in humans and animals shows that task proficiency often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. However, these results are derived from models that assume the same increase in expertise for each individual, and the data from these studies does not include repeated measures of individual success through time. In contrast, our analyses revealed that there was no population increase in expertise; instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition. For instance, many predators increased their success (Figure S1A), but others displayed a decrease in success (Figure S1B) or no change in success at all with experience. We predicted that this would occur because predators should differ in the type of prey that they encounter, as some may be easier to capture than others. This is often the case in nature as prey can use variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. However, prey speed in our study did not shape individual differences in expertise acquisition by predators, even if predators differed in the prey that they encountered. This was unexpected, because prey speed is an important predictor of predator success in this system as well as in others [@Walker.etal2005;@Kelley.Magurran2011;@FraserFranco.etal2022]. Despite that other antipredator tactics may mediate these individual responses by predators, it is likely that individual differences in motivation and perseverance, attention span, or memory retention were also major causes [@Warburton2003;@morand-ferronStudyingEvolutionaryEcology2016;@Dukas2019]. For instance, a persevering predator may increase its success much more steadily than an unmotivated predator that quickly abandons a chase. Another factor that may explain differences in the development of expertise is the time delay between hunting events. For example, a predator that played 300 matches in the span of six months might forget more critical information (e.g. prey escape patterns or muscle memory) than one that played 300 matches in the span of 6 days. While this has not been formally tested, a greater delay between exposure events is hypothesized to attenuate or even negate the relationship between cumulative experience and success [@Endler1991;@Wright.etal2022].

>>>>

---
title: "Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame"
output: 
  word_document:
    reference_docx: Chapter2-styles.docx
    fig_caption: yes
    pandoc_args: ["-Fpandoc-crossref"] # to have eq. numbers
bibliography: references2.bib
csl: chicago-author-date.csl
---

@@

We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used Gaussian priors for the game duration ($N(0,\ 1)$), the intercept ($N(0,\ 0.5)$), and the precision paremeter ($N(2,\ 0.5)$). We employed the default Student t priors for the smoothing parameter ($t(3,\ 0,\ 2.5)$). In all models, we specified four MCMC chains to estimate the parameters, with 1000 posterior samples for each parameter. To do so, we ran 2500 iterations for the first model with a thinning set to eight, and 1500 iterations with a thinning set to four for the second and third models; burn-in was set to 500 iterations in each model. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021a]. After fitting the three models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

We parameterized equations (1) and (2) with a Gaussian distribution family (predator and prey speed) and used a beta-binomial distribution family for equation (3) (predator hunting success). We used Gaussian priors for the prey rank ($N(2.5,\ 1)$) and game duration ($N(10,\ 2)$) on the mean part of the model, and the default priors in "brms" for the dispersion part. We used the default Student t priors in "brms" for the intercepts on the mean and dispersion part of the equation. For hunting success, we applied a Gaussian prior on the precision parameter ($N(2,\ 1)$). For every predicted variable, we applied a half-Gaussian on the random effects for the mean part of the model ($N(0,\ 1)$), and the default "brms" Student t prior for the dispersion part of the model. We parametrized the model to run four MCMC chains with 1000 posterior samples for each parameter by completing 2500 iterations having a thinning set to 8, with the first 500 iterations used as warmups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021a]. We also performed posterior predictive checks which showed an adequate fit of the model (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We additionally report variance partitioning and coefficients of variations in the supporting information (Appendix S1).

@@

Our results confirm that individual predators display substantial differences in the development of their expertise, as the best model included group-level smoothers exclusively (Table I). For instance, some predators steadily increased (Figure 1A, Appendix S1: Figure S1A), while others decreased (Figure 1A, Appendix S1: Figure S1B), their hunting success with greater experience, and others whose success did not change with experience (Figure 1A). Another pattern that emerged was that some individuals optimized their success at an experience level that was below what we considered "advanced" (i.e. >300 matches played).

@@

At the population level, we predicted that predators would become either increasingly flexible or specialized with more experience. However, our results do not support this prediction. Indeed, the observed pattern is more complex than expected given that predators became more flexible with experience at intermediate levels of experience (i.e. intercept of sigma for predator speed is larger for intermediates than novices; Table 2 and Appendix S1: Figure S4), but then return to novice-levels with advanced experience (i.e. intercept of sigma for predator speed is smaller for advanced players than intermediates; Table 2 and Appendix S1: Figure S4). Thus, our analyses suggest that population-level behavioural specialization changes nonlinearly (quadratic) with experience rather than linearly.

@@

At the individual level, we detected differences in the mean movement speed among predators (i.e. the standard deviation of the mean speed for the predator ID; Table 3). These differences largely remained stable to the intermediate stage and then increased at the advanced stage to a level higher than when predators were novice (Table 3 and Appendix S1: Figure S4). For the dispersion part of the model, predators displayed considerable differences in their degree of hunting specialization, which also increased with experience (i.e. the standard deviation of sigma for the predator ID increases with experience; Table 3 and Appendix S1: Figure S4). Thus, the predator population is composed of both specialist and flexible individuals, and they further distinguish themselves as specialist/flexible hunters as they gain experience (Figure 2).

@@

As we predicted, predators differed in the mean speed of the prey that they encountered. Thus, some individuals encountered prey that were on average slower, while others encountered prey that where on average faster throughout the study period (i.e. the standard deviation of the mean prey speed for the predator ID; Table 3). These differences were greater when predators reached an advanced level of hunting experience (i.e. the standard deviation of the mean prey speed for the predator ID was higher at the advanced stage; Table 3 and Appendix S1: Figure S4). In addition, individuals experienced largely distinct degrees of variability among their encounters with the prey. Hence, some predators experienced stable/predictable encounters with their prey, while others experienced unstable/unpredictable encounters with their prey throughout the study period (i.e. the standard deviation of sigma prey speed for the predator ID; Table 3).

@@

At all experience levels, predators that were on average faster (mean) also had a lower intra-individual variability (IIV) across their matches (Figure 3 and Appendix S2: Table S1). Thus, faster predators specialized at travelling fast, while slower predators were more flexible in their movement through time. In general, predators that encountered prey that were on average faster (mean) tended to be faster (mean) (Figure 3 and Appendix S2: Table S1). At the three experience levels, there were only weak correlations between predator IIV and mean prey speed (Figure 3 and Appendix S2: Table S1). However, predators that experienced more predictable encounters with their prey (IIV) specialized in a faster hunting tactic (IIV) (Figure 3 and Appendix S2: Table S1). For all the results described above, the correlations strengthen as predators gain experience (Figure 3 and Appendix S2: Table S1).

The foraging strategies of predators, that is, the correlations between the mean and IIV of speed, remained structurally stable among the levels of experience (Figure 3 inside the black squares and Appendix S2: Table S1). For example, novice predators that tended to be faster maintained this strategy as they gained experience (Figure 3 inside the black squares and Appendix S2: Table S1). Overall, we detected the same patterns as those from the within-experience state correlations described above. The correlations between trait values were also stronger among novice-intermediate and intermediate-advanced states than between novice-advanced states, indicating that the hunting strategies of predators were more similar among experience levels closer in time (Figure 3 inside the black squares and Appendix S2: Table S1).

@@

We did not detect strong correlations between the predators' mean speed and their hunting success (Figure 3 and Appendix S2: Table S1), nor between the predators' IIV in speed and their hunting success (Figure 3 and Appendix S2: Table S1), suggesting that slower/flexible and faster/specialist predators were both equally successful. However, when predators were novice, there was a strong negative correlation between the mean speed of the prey that they encountered and their hunting success (Figure 3 and Appendix S2: Table S1). Hence, novice individuals that encountered faster prey were less successful. This relationship was decoupled at the later experience levels, such that encounters with faster prey at intermediate and advanced stages were not as strongly correlated with hunting success anymore (Figure 3 and Appendix S2: Table S1). Thus, the consequences of hunting faster prey were reduced at greater experience levels.

@@

Empirical research in humans and animals shows that task proficiency often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. However, these results are derived from models that assume the same increase in expertise for each individual, and the data from these studies does not include repeated measures of individual success through time. In contrast, our analyses revealed that there was no population increase in expertise; instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition. For instance, many predators increased their success (Appendix S1: Figure S1A), but others displayed a decrease in success (Appendix S1: Figure S1B) or no change in success at all with experience. We predicted that this would occur because predators should differ in the type of prey that they encounter, as some may be easier to capture than others. This is often the case in nature as prey can use variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. However, prey speed in our study did not shape individual differences in expertise acquisition by predators, even if predators differed in the prey that they encountered. This was unexpected, because prey speed is an important predictor of predator success in this system as well as in others [@Walker.etal2005;@Kelley.Magurran2011;@FraserFranco.etal2022]. Despite that other antipredator tactics may mediate these individual responses by predators, it is likely that individual differences in motivation and perseverance, attention span, or memory retention were also major causes [@Warburton2003;@morand-ferronStudyingEvolutionaryEcology2016;@Dukas2019]. For instance, a persevering predator may increase its success much more steadily than an unmotivated predator that quickly abandons a chase. Another factor that may explain differences in the development of expertise is the time delay between hunting events. For example, a predator that played 300 matches in the span of six months might forget more critical information (e.g. prey escape patterns or muscle memory) than one that played 300 matches in the span of 6 days. While this has not been formally tested, a greater delay between exposure events is hypothesized to attenuate or even negate the relationship between cumulative experience and success [@Endler1991;@Wright.etal2022].
